Web Authn Reference Implementation

### 2019-02-04 doc/paper/paper.tex
draft 2

% Glossary
\usepackage{glossaries}
\makeglossaries
\input{glossary.tex}

@@

Passwords are the most common used way to authenticate a user. They are easy to implement and have been used since the beginning of the internet. People are used to passwords and use them everywhere, but most of them use weak ones. Remembering strong passwords is not easy, thus, most opt for easier passwords or password reuse them. The newest recommendations from the \gls{nist} are actually to use very long passwords that make sense for a human. For instance a sentence in a fictional language or with slang words. An appropriate password would then look something like `awsumfurrycybercat', but not that. \cite{nist:pw:blog}

Even when user use strong passwords with this it is hard to remember a new password for every site. Thus, they will reuse passwords for multiple sites. Password reuse can lead to dangerous password credential stuffing attacks where the password that was leaked from one service can be used to access another service. \cite{panda:pwreuse, xkcd:pwreuse} Those are two weaknesses that passwords always had. Firstly, used passwords are generally too weak, secondly, they are reused for multiple services.

\gls{fido2} is a specification that aims to increase security for end users by eliminating the weaknesses that come with passwords. The main advantages are strong privacy, privacy protection, multiple choices, cost-efficiency and a layered approach. It tries to do that by using secure authenticators that are used instead of a password. \cite{yubico:whatIsFido2}

The main force behind \gls{fido2} is the \gls{fido} Alliance. Some noticable members are Microsoft, Google and Yubico. Microsoft for instance wants to use \gls{fido2} for the windows hello login. \cite{yubico:ms} With that force behind it it has been adopted in all major web browser. \cite{fido:browser} There are also already authenticators available from yubico that can be used. \cite{yubico:yubikey5}

@@

\gls{fido2} is an authentication standart wich uses public key cryptography to authenticate an user. Compared to a password the user doesn't prove that he knows something but that he has something that he previously has registered with the application. This authenticator is an external device that interacts with the application through a protocol called \gls{ctap}. \gls{ctap} is also part of the \gls{fido2} specification but this paper does not go into detail of how this protocol works. \cite{ctap}

The \gls{webAuthn} specification forms the second part of \gls{fido2}. This part deals with the Application that wants to authenticate an user. It contains two steps. Attestation which deals with registering a new user and assertion which handles the authentication of an existing user. Both functions are explained in the Section \ref{sec:replying_party}.

@@

Traditionally, when a user wants to register or authenticate, he supplies the application with a password. A shared secret that only the user is supposed to know. We already discussed how passwords usually lead to a weak authentication. This issue was usually solved by adding a second factor to the authentication. With \gls{fido} \gls{u2f} this second factor would be very strong and the authentication would be resilient against scaled attacks. However, many sites offer weaker forms of second factor authentication like codes via e-mail or phone. Both of those methods are weak to message interception where an attacker intercepts the message before the end user recieves it. \cite{smsweak}

By using certified authenticators, \gls{fido2} makes sure that a user needs to have access to his authenticator and be present when the authentication is triggered. With this an attacker could still steal the physical authenticator and authenticate with it. However, this kind of physical attack can't be used to steal the authenticator for a large group of people. Additionally once the user notices that his authenticator has been stolen, he can inform the service provider and disable this authenticator for future authentications. Additionally to secure authenticators that are simply left in the USB slot of a machine, \gls{fido2} requires the user to be present during the authentication. With this an attacker can't simply put some malware on the users computer and trigger the login.

\subsection{Authenticator}
\gls{fido2} authenticators are certified devices that support the \gls{fido2} standart. They must at least fulfill level 1 of the certification levels. There are 3 levels that can be achieved that all build on each other plus one for additional security. The levels are described below. New authenticators need to first get a functional certificate to profe that they are interoperable with other clients and servers.

The authenticator used in this reference implementation is a Yubikey 5 which supports an interface for FIDO2 via \gls{usb} and via \gls{nfc}. In this paper only the \gls{usb} interface will be used.

@@

This level needs an authenticator to run in a \gls{fido} \gls{aroe} and needs to support allowed cryptography lists. Examples that would not achieve this level are: \cite{fido:authenticator:level2, fido:authenticator:allowedEnvironment}
\begin{enumerate}
  \item Authenticator that doesn't support attestation
  \item Authenticator that is an implementation without restricted operating environment 
\end{enumerate}

@@

\gls{fido} \gls{u2f} is the predecessor of \gls{fido2}. Where \gls{fido2} covers more usecases than \gls{fido} \gls{u2f}. The latter is only for second factor authentication. \gls{fido2} can be used as a second factor but also for passwordless authentication and local multifactor authentication. \cite{yubico:whatIsFido2}

It is possible to use a \gls{fido} \gls{u2f} token to authenticate with a \gls{fido2} replying party and it is also possible to use a \gls{fido2} authenticator with a \gls{fido} \gls{u2f} implementation. This mapping is mainly done in the \gls{ctap} version 1 for \gls{fido} \gls{u2f} and version 2 for \gls{fido2}. In the \gls{fido2} version there is a mapping defined which can map a response of a \gls{fido} \gls{u2f} authenticator to a response of a \gls{fido2} authenticator. This mapping for the authentication can be seen in Figure \ref{fig:authenticationMapping}. The full description is available in the \gls{ctap} specification. \cite{yubico:whatIsFido2, ctap:interoperability}

\begin{figure}[ht]
  \includegraphics[width=12cm]{img/u2fcompat-getassertion.png}
  \centering
  \caption{Authenticator Data}
  \label{fig:authenticationMapping}
\end{figure}

@@

The replying party is one of the main components of the \gls{fido2} standart and the main component of this paper. It includes a sample implementation of the replying party meant as a reference. This part will deal with the two phases attestation and authentication. First there is a short section dealing with the setup of the reference code.

@@

The reference code is written in Java and uses Maven for dependency management. To get the \gls{rest} \gls{api} set up Spring Boot is used such that most of the non relevant code of setting up controllers etc could be spared. The data from the registrations are stored in a set to decrease the non-relevant code even more. There are some data classes that mostly represent data that is used within the application.

The frontend part is from the \gls{fido} Alliance demo \cite{fido:demo}. Additionally some inspiration was taken from the Yubico implementation of the Java-Webauthn-Server also available on Github \cite{yubico:webauthn:server}.

For testing the YubiKey 5 was used \cite{yubico:yubikey5} which is reflected in the code. There are multiple ways with which the the standart can be implemented. Only the way to implement it with a yubikey 5 is available in the reference code. The other ways are discussed in their relative paragraphs.

The relevant classes are the LoginController, the RegistrationController and the AuthenticatorDataParser. Those are nice references for when one wants to implement the replying party oneself. Those classes are displayed in depth in the following sections.

@@

The registration which is reflected in the RegistrationController represents the process a user has to go through to register with the application using a \gls{fido2} authenticator. It consists of two main steps. First the replying party has to create a \gls{json} that will be accepted by the token in the \gls{ctap} protocol as a valid challenge. After the response from the client the response endpoint need to validate the input from the authenticator. This can be seen in Figure \ref{fig:user_registration}.

@@

The create registration step consists of creating a \gls{json} response that the autenticator can reply to. This JSON is sent to the browser where it calls `navigator.create()'. The respons is then sent to the /response endpoint in the replying party.

@@

\paragraph{Step1} \hfill \\ 
In Java with spring boot step 1 is mostly not needed. But checking that the needed fields in the json are present has to be done at some point.
\paragraph{Step2} \hfill \\ 
In step 2 the clientDataJSON field has to be parsed. This is done by base 64 url decoding the clientDataJSON field and then representing the data as an \gls{utf-8} string which can be parsed into \gls{json}. Parsed it looks like in Listing \ref{lst:clientDataParsed}

@@

\paragraph{Step4}\hfill \\ 
In step 4 we need to verify the challenge. In the code this is done by searching for the user with this challenge. If none is present or the user is already registered an error code is returned.

\paragraph{Step5}\hfill \\ 
Step 5 is trivial. The origin that is returned must match the origin of the replying party. For convenience in the code just the presence of the origin is checked. It is important that in a productive environment this doesn't suffice. This string needs to match the URL where it was called. This check offers protection against \gls{mitm} attacks.

@@

\paragraph{Step7}\hfill \\ 
In step 7 the \gls{sha-256} hash for the signature is precomputed. It is important that the string value of the json property `clientDataJSON' is compared, else the signature will be incorrect.

\paragraph{Step8}\hfill \\ 
Step 8 is a lot of work. First the `attestationObject' from the `response' has to be base64 url decoded and then parsed into \gls{json} using a \gls{cbor} mapper. The result of this is shown in Listing \ref{lst:attestationData}. After that the `authData' field of the attestation data needs to be parsed. It is encoded in a specified format. The decoding of this is described in Section \ref{sec:authData}. \cite{webauthn:authData}

@@

\paragraph{Step9}\hfill \\ 
In step 9 the `rpIdHash' from the decoded auth data is compared to the \gls{sha-256} hash of the rp.id from the Listing \ref{lst:createRegistration}. This is used to verify again that the token has registered with the correct replying party. Even though this was tested already it should be checked again to protect against some forms of attack like \gls{mitm}.

\paragraph{Step10}\hfill \\ 
In step 10 the user present flag has to be verified. This is one of the main features of \gls{fido2} as it ensures that an user was present when the attestation was created. However, this doesn't prove that the user was the user that he claims to be, but simply that he is a user. This limits many kinds of remote attacks as they can't trigger the authenticator to set this flag to true. Physical attacks are still possible though.

\paragraph{Step11}\hfill \\ 
In step 11 the user verified flag is checked. This is optional and doesn't necessarily need to be checked. Only if the authenticator needs to support it. If this is required it is advisable to inform the user beforehand and if the bit is not set, he should get informed that he should use a different token.

@@

\paragraph{Step13}\hfill \\ 
Step 13 verifies the `fmt' statement. As the implementation only validates yubikey5 authenticators, it is set to the value that is given back by those which is packed.

\paragraph{Step14}\hfill \\ 
In step 14 the signature is verified. For this there are multiple paths defined by the format. With the packed format there are multiple options of which yubikey 5 uses the x509 cert. This certificate first needs to be parsed. After it has been parsed some attributes must be checked. Only after those steps the signature is validated. This is done by first combining the authenticator data with the hash computed at step 7.

\paragraph{Step15}\hfill \\ 
Step 15 validates the trust anchor, in other words, whether we should trust this authenticator or not. This should be done by getting data from a fido metadata service. In the implementation it is done by comparing the aaguid from the authenticator to the one recieved from a Yubikey 5. This should be the same for each key.

\paragraph{Step16}\hfill \\ 
In step 16 the trustworthiness of the certificate used above should be checked. Again as we only trust yubico authenticators the ca cert from yubico is staticly used.

\paragraph{Step17}\hfill \\ 
Step 17 is testing for same credential on multiple users. The behavior of this is not specific. Additionally while testing the same authenticator always returned a different credential id and public key. Still, the check is done by checking against the other users.

\paragraph{Step18}\hfill \\ 
In step 18 the user is associated with the credential. It is added to that user and used later for a authentication. The state of the user is also set to registered.

\paragraph{Step19}\hfill \\ 
Step 19 only states that if step 16 failed, the ceremony should be failed. As this was already done at step 16 this step can be ignored.

\subsubsection{Summary}
Overall there are a lot of steps needed for a successful ceremony. None of the steps should be skipped for convenience as all of them provide some sort of protection against various forms of attack. Some checks are rather complex and a real implementation should be wary of that. Maybe covering only one path is the simplest solution and fulfills all the requirements for a specific application. If that is the case only this way should be implemented. More complexity often only increases risk of exposure.

@@

The authentication which is implemented in the LoginController deals with the process of authenticating to a replying party by using an authenticator that has previously been registered. There are again two main stages. The replying party again has to first create a \gls{json} which contains a challenge. This challenge is responded by the authenticator and in a second step the replying party has to verify that response by checking against the authenticator data that were registered before. The process is depicted in Figure \ref{fig:user_authentication}.

@@

\begin{description}
  \item[challenge] \hfill \\ The challenge to be signed by the Authenticator. Must be unique for each call and base 64 encoded.
  \item[fidoResponse] Should be direct. 
  \item[rpid] \hfill \\ Must reflect the domain over which you communicate
  \item[allowCredentials.type] Should be public-key. 
  \item[allowCredentials.id] Should be the id the user used when registering.
  \item[timeout] Should be defined.   
\end{description}

@@

To verify an authentication assertion the standart again supplies a set of steps. This time there are 18. Some steps are not represented directly in the implementation. This is mostly because they can't be represented directly. Additionally before the first step some serialization has to be done. The initial response is visible in Listing \ref{lst:attestationResponse}.

@@

\paragraph{Step1}\hfill \\ 
In the first step the credential should be matched to the credentials requested. This is done by first getting the user by the used credential. Then the stored assertion request for this user is compared against the credential id given with the assertion.

\paragraph{Step2}\hfill \\ 
The user has already been retrieved in step 1. In step 2 the userhandle should be verified. The userhandle is optional but if it is present it must reflect the user that was challenged.

\paragraph{Step3}\hfill \\ 
In step 3 the public key must be retrieved. This is stored with the user.

@@

\paragraph{Step6}\hfill \\ 
Step 6 requires a \gls{json} parsing of the decoded cData. Step 5 and 6 are to close together that the reference implementation combined them into one step which deals with both at the same time. The final clientDataJSON can is displayed in Listing \ref{lst:attestationClientData}.

@@

\paragraph{Step9}\hfill \\ 
In step 9 the origin has to be verified. Again as in the registration process this is done by checking if the origin contains the domain. In a productive environment this should be done by comparing the expected uri.

\paragraph{Step10}\hfill \\ 
Step 10 deals with token binding which is optional and currently not supported in the reference implementation.

\paragraph{Step11}\hfill \\ 
In step 11 the rpIdHash from the authData must equal to the \gls{sha-256} hash of the RP ID. For this first the authData needs to be parsed.

\paragraph{Step12}\hfill \\ 
In step 12 the user present flag is checked again. This is the same process as in the registration and must be checked to protect against remote attacks.

@@

\paragraph{Step15}\hfill \\ 
In step 15 the client data is hashed using \gls{sha-256}. This needs to be the exact string that was recieved by accessing the clientDataJSON field.

@@

\paragraph{Step17}\hfill \\ 
Step 17 is needed to check for cloned authenticators. If the signCount is present it should always increase. It is important to verify this behavior to avoid cloned authenticators.

\paragraph{Step18}\hfill \\ 
Step 18 states that should any step above have been invalid, the authentication ceremony must be failed.

\subsubsection{Summary}
After already knowing what happens in the registration the authentication is very simmilar. Many steps exist in both ceremonies in the exact same or a very simmilar way. The steps of the assertion sometimes can't be represented exactly in code because with java it is sometimes easier to combine two steps into one. However, it is important that all of the steps are followed as required. Missing some steps could lead to a broken authentication.

\subsection{Auth Data}
\label{sec:authData}
In both the registration and authentication ceremony there is one step where the authenticator data must be parsed. Thus this was done at a central place called the `AuthenticatorDataParser'. This data comes in binary and needs to be parsed according to the specification. The first 32 bytes generate the rpIdHash, the next byte the flags sector. Then there are 4 bytes which represent the signature Counter. After that there might exist some attestedCredentialData. A table with all the information can be seen in Figure \ref{fig:authenticatorData}.

This attestedCredentialData if present contains 16 bytes of the aaguid, with which the authenticator model can be determinded, 2 bytes which set the lenght of the following field which is the credential id and the rest which makes the credential Public Key. This attestedCredentialData is present for registration to verify the assertion. A table with all the information can be seen in Figure \ref{fig:credentialData}.

@@

\section{Conclusion}
The \gls{fido2} standart increases the security of the end user by eliminating passwords with a \gls{mitm} resistant factor. The authenticator can be supplied by many parties which makes wide adoption easier. However, currently the implementation of the standart for a webpage owner is probably to complex. There are not yet any libraries for all the languages available. This needs to change if wide adoption should be achieved. Additionally, there are not yet a lot of \gls{fido2} authenticators available on the market. Which results in a poor adoption rate.

However, the standart is quite new and could be adopted widely. It has a strong support from Microsoft, where it is implemented in Windows Hello. This could result in many companies supplying authenticators to their employees which would lead to a higher incentive to provide web applications with \gls{fido2} support. It is important that other big players implement the standart so that it truly can replace passwords.

An interesting application of this would be within companies to replace the need for passwords. Also it could be used by governments in combination with an \gls{eid} to give access to certain resources by using a phyiscal card as an authenticator. This way every citizen of a country already has access to a \gls{fido} certified authenticator which would lead to country specific sites implementing the standart as an authentication method.

@@

This implementation only covered the yubikey 5 authenticator. It would be interesting to see how other authenticators react. Also it would be interesting to see how the yubikey 5 interacts via the \gls{nfc} interface. This implementation was only tested using the \gls{usb} interface of the authenticator.

Additionally to that it only dealt with authentication and registration. It is important to give the user the option to add multiple authenticators to his account such that he can recover his account should he loose his key. Additionally it is important that the user can revoke their authenticators. This should be done by giving the user the option to set a name that is relevant to him for each authenticator. He should then be able to manage those authenticators. Either adding new ones or removing existing ones.

>>>>

% Glossary
\usepackage[toc,section=section]{glossaries}
\makeglossaries
\input{glossary.tex}

@@

Passwords are the most commonly used way to authenticate a user. They are easy to implement and have been used since the beginning of the internet. People are used to passwords and use them everywhere, but most of them use weak ones. Remembering strong passwords is not easy; thus, most opt for easier passwords or password reuse them. The newest recommendations from the \gls{nist} are actually to use very long passwords that make sense for a human — for instance a sentence in a fictional language or with slang words. An appropriate password would then look something like `awsumfurrycybercat', but not that. \cite{nist:pw:blog}

Even when the user employ strong passwords, it is hard to remember a new password for every site. Thus, they reuse passwords for multiple sites. Password reuse can lead to \gls{csa}. \cite{panda:pwreuse, xkcd:pwreuse} Those are two weaknesses that passwords always had. Firstly, used passwords are generally too weak; secondly, they are reused for multiple services.

\gls{fido2} is a specification that aims to increase security for end users by eliminating the weaknesses that come with passwords. The main advantages are strong privacy, privacy protection, multiple choices, cost-efficiency, and a layered approach. It tries to do that by using secure authenticators that are used instead of a password.  Additionally to the already discussed attacks \gls{fido2} also protects against \gls{mitm} attacks. \cite{yubico:whatIsFido2}

The main force behind \gls{fido2} is the \gls{fido} Alliance. Some notable members are Microsoft, Google, and Yubico. Microsoft, for instance, wants to use \gls{fido2} for the windows hello login. \cite{yubico:ms} With that force behind the standard has is implemented in all major web browser. \cite{fido:browser} There are also already authenticators available from Yubico that can support it. \cite{yubico:yubikey5}

@@

\gls{fido2} is an authentication standard which uses public key cryptography to authenticate the user. Compared to a password the user doesn't prove that he knows something but that he has something that he previously has registered with the application. It is also possible to secure this authenticator with a second factor like a pin or biometrics. The authenticator is an external device that interacts with the application through a protocol called \gls{ctap}. \gls{ctap} is also part of the \gls{fido2} specification but this paper does not go into detail about how this protocol works. \cite{ctap}

The \gls{webAuthn} specification forms the second part of \gls{fido2}. This part deals with the Application that wants to authenticate the user. It contains two steps. Attestation which deals with registering a new user and assertion which handles the authentication of an existing user. Both functions are explained in the Section \ref{sec:replying_party}.

@@

Traditionally, when a user wants to register or authenticate, he supplies the application with a password. A shared secret that only the user is supposed to know. We already discussed how passwords usually lead to weak authentication. This issue is normally circumvented by adding a second factor to the authentication. With \gls{fido} \gls{u2f} this second factor would be robust, and the authentication would be resilient against scaled attacks. However, many sites offer weaker forms of second-factor authentication like codes via e-mail or SMS. Both of those methods are weak to message interception. \cite{smsweak}

By using certified authenticators, \gls{fido2} makes sure that a user needs to have access to his authenticator and be present when the authentication is triggered. With this, an attacker could still steal the physical authenticator and authenticate with it. However, this kind of physical attack can't be used to steal the authenticator for a large group of people. Additionally, once the user notices that his authenticator has been stolen, he can inform the service provider and disable this authenticator for future authentications. Moreover, to secure authenticators that are left in the USB slot of a machine, \gls{fido2} requires the user to be present during the authentication. With this, an attacker can't merely put some malware on the user's computer and trigger the login.

\subsection{Authenticator}
\gls{fido2} authenticators are certified devices that support the \gls{fido2} standard. They must at least fulfill level 1 of the certification levels. 3 levels can be achieved that all build on each other plus one for additional security. New authenticators need first to get a functional certificate to profe that they are interoperable with other clients and servers.

The authenticator used in this reference implementation is a Yubikey 5 which supports an interface for \gls{fido2} via \gls{usb} and via \gls{nfc}. In this paper, only the \gls{usb} interface is used.

@@

This level needs an authenticator to run in a \gls{fido} \gls{aroe} and needs to support allowed cryptography lists. Examples that would not achieve this level are: \cite{fido:authenticator:level2, fido:authenticator:allowedEnvironment}
\begin{enumerate}
  \item Authenticator that doesn't support attestation
  \item Authenticator that is an implementation without a restricted operating environment 
\end{enumerate}

@@

\gls{fido} \gls{u2f} is the predecessor of \gls{fido2}. Where \gls{fido2} covers more usecases than \gls{fido} \gls{u2f}. Where \gls{fido} \gls{u2f} is only for second factor authentication, \gls{fido2} can be used as a second factor but also for passwordless authentication and local multifactor authentication. \cite{yubico:whatIsFido2}

It is possible to use a \gls{fido} \gls{u2f} token to authenticate with a \gls{fido2} replying party and vice verca. This mapping is mainly done in the \gls{ctap} version 1 for \gls{fido} \gls{u2f} and version 2 for \gls{fido2}. In the \gls{fido2} version there is a mapping defined which can map a response of a \gls{fido} \gls{u2f} authenticator to a response of a \gls{fido2} authenticator. This mapping for the authentication can be seen in Figure \ref{fig:authenticationMapping}. The full description is available in the \gls{ctap} specification. \cite{yubico:whatIsFido2, ctap:interoperability}

\begin{figure}[h]
  \includegraphics[width=12cm]{img/u2fcompat-getassertion.png}
  \centering
  \caption{Authenticator Data}
  \label{fig:authenticationMapping}
\end{figure}

@@

The replying party is one of the main components of the \gls{webAuthn} standard and the main component of this paper. It includes a sample implementation of the replying party meant as a reference. This part deals with the two phases attestation and authentication. First, there is a short section dealing with the setup of the reference code.

@@

The reference code is written in Java and uses Maven for dependency management. To get the \gls{rest} \gls{api} set up Spring Boot is used such that most of the non-relevant code isn't needed. The data from the registrations are stored in a set to decrease the non-relevant code even more. Additionally, there are some data classes which are used within the application.

The frontend part is from the \gls{fido} Alliance demo \cite{fido:demo}. Additionally, some inspiration was taken from the Yubico implementation of the Java-Webauthn-Server also available on Github \cite{yubico:webauthn:server}.

For testing the YubiKey 5 was used \cite{yubico:yubikey5} which is reflected in the code. There are multiple ways with which the standard can be implemented. Only the way to implement it with a Yubikey 5 is available in the reference code. The other ways are discussed in their paragraphs.

The most important classes are the LoginController, the RegistrationController, and the AuthenticatorDataParser. Those are supposed to be references for when one wants to implement the replying party oneself. Those classes are discussed in depth in the following sections.

For testing the replying party it is important to generate a valid certificate and to communicate with a valid domain. If the replying party listens to localhost the response is invalid. The same happens when the communication is not secured with a valid certificate and in a \gls{tls} connection.

@@

The registration which is reflected in the RegistrationController represents the process a user has to go through to register with the application using a \gls{fido2} authenticator. It consists of two main steps. First, the replying party has to create a \gls{json} that is accepted by the token in the \gls{ctap} protocol as a valid challenge. After the response from the client, the response endpoint needs to validate the input from the authenticator. This can be seen in Figure \ref{fig:user_registration}.

@@

The create registration step consists of creating a \gls{json} which represents a valid challenge. This JSON is sent to the browser where it is handed to the call `navigator.create()'. The response is sent to the /response endpoint in the replying party.

@@

\paragraph{Step1} \hfill \\ 
In Java with Spring Boot, the first step is mostly not needed. But the check that the needed fields in the \gls{json} are present can be done at this point.

\paragraph{Step2} \hfill \\ 
In step 2 the clientDataJSON field has to be parsed. This is done by \gls{base64} URL decoding the clientDataJSON field and then representing the data as an \gls{utf-8} string which can be parsed into \gls{json}. Parsed it looks like in Listing \ref{lst:clientDataParsed}

@@

\paragraph{Step4}\hfill \\ 
In step 4 we need to verify the challenge. In the code, this is done by searching for the user with this challenge.

\paragraph{Step5}\hfill \\ 
Step 5 is trivial. The origin that is returned must match the origin of the replying party. For convenience in the code, just the presence of the origin is checked. It is important that in a production environment this doesn't suffice. This string needs to match the URL where it was called. This check offers protection against \gls{mitm} attacks.

@@

\paragraph{Step7}\hfill \\ 
In step 7 the \gls{sha-256} hash for the signature is precomputed. It is important that the string value of the \gls{json} property `clientDataJSON' is compared, else the signature is incorrect.

\paragraph{Step8}\hfill \\ 
Step 8 needs much work. First the `attestationObject' from the `response' has to be \gls{base64} URL decoded and then parsed into \gls{json} using a \gls{cbor} mapper. The result of this is shown in Listing \ref{lst:attestationData}. After that, the `authData' field of the attestation data needs to be parsed. It is encoded in a specified format. The decoding of this is described in Section \ref{sec:authData}. \cite{webauthn:authData}

@@

\paragraph{Step9}\hfill \\ 
In step 9 the `rpIdHash' from the decoded auth data is compared to the \gls{sha-256} hash of the rp.id from the Listing \ref{lst:createRegistration}. This process is used to verify again that the token has registered with the correct replying party. Even though this was tested already, it should be rechecked to protect against some forms of attack like \gls{mitm}.

\paragraph{Step10}\hfill \\ 
In step 10 the user present flag has to be verified. This process is one of the main features of \gls{fido2} as it ensures that a user was present when the attestation was created. However, this doesn't prove that the user was the user that he claims to be, but simply that he is a user. This step limits many kinds of remote attacks as they can't trigger the authenticator to set this flag to true. Physical attacks are still possible though.

\paragraph{Step11}\hfill \\ 
In step 11 the user verified flag is checked. This verification is optional and doesn't necessarily need to be checked. Only if the authenticator needs to support it. If this is required, it is advisable to inform the user beforehand, and if the bit is not set, he should get informed that he should use a different token or that he needs to be verified on authentication.

@@

\paragraph{Step13}\hfill \\ 
Step 13 verifies the `fmt' statement. As the implementation only validates Yubikey 5 authenticators, it is set to the value that is given back by those which is packed.

\paragraph{Step14}\hfill \\ 
In step 14 the signature is verified. For this, there are multiple paths defined by the format. With the packed format, there are multiple options of which Yubikey 5 uses the x509 cert. This certificate first needs to be parsed. After it has been parsed some attributes must be checked. Only after those steps, the signature is validated. This validation is done by first combining the authenticator data with the hash computed at step 7 and test this data against the signature value provided.

\paragraph{Step15}\hfill \\ 
Step 15 validates the trust anchor, in other words, whether we should trust this authenticator or not. This process should be done by getting data from a \gls{fido} metadata service. In the implementation, it is done by comparing the `aaguid' from the authenticator to the one received from a Yubikey 5. This guid should be the same for each key of the same model.

\paragraph{Step16}\hfill \\ 
In step 16 the trustworthiness of the certificate used above should be checked. Again as we only trust Yubico authenticators, the CA cert from Yubico is statically used.

\paragraph{Step17}\hfill \\ 
Step 17 is testing for the same credential on multiple users. The behavior of this is not specific. Additionally while testing the same authenticator always returned a different credential id and public key. Still, the check is done by checking against the other users.

\paragraph{Step18}\hfill \\ 
In step 18 the user is associated with the credential. It is added to that user and used later for an authentication. The state of the user is also set to register.

\paragraph{Step19}\hfill \\ 
Step 19 only states that if step 16 failed, the ceremony should be failed. As this was already done at step 16, this step can be ignored.

\subsubsection{Summary}
Overall there are many steps needed for a successful ceremony. None of the steps should be skipped for convenience as all of them provide some protection against various forms of attack. Some checks are rather complex, and a real implementation should be wary of that. Maybe covering only one path is the simplest solution and fulfills all the requirements for a specific application. If that is the case only this way should be implemented. More complexity often only increases the risk of exposure.

@@

The authentication is implemented in the LoginController. It deals with the process of authentication to a replying party by using an authenticator that has previously been registered. As with the registration, there are two main stages. The replying party again has first to create a \gls{json} which contains a challenge. , and in a second step, the replying party has to verify that response by checking against the authenticator data that were registered before. The process is depicted in Figure \ref{fig:user_authentication}.

@@

\begin{description}
  \item[challenge] \hfill \\ The challenge to be signed by the Authenticator. Must be unique for each call and \gls{base64} encoded.
  \item[fidoResponse] Should be direct. 
  \item[rpid] \hfill \\ Must reflect the domain over which you communicate
  \item[allowCredentials.type] Should be public-key. 
  \item[allowCredentials.id] Should be the id the user used when registering.
  \item[timeout] Should be defined.   
\end{description}

@@

The standard supplies a set of 18 steps to verify an authentication assertion. Should a step fail, a LoginFailedException is thrown which results in an error code in return. Before the first step, some serialization has to be done. The initial response is visible in Listing \ref{lst:attestationResponse}.

@@

\paragraph{Step1}\hfill \\ 
In the first step, the credential should be matched to the credentials requested. This match is done by first getting the user with the credential. Then the stored assertion request for this user is compared against the credential id given with the assertion.

\paragraph{Step2}\hfill \\ 
The user has already been retrieved in step 1. In step 2 the user handle should be verified. The user handle is optional, but if it is present, it must reflect the user that was requested.

\paragraph{Step3}\hfill \\ 
In step 3 the public key must be retrieved. It is stored with the user.

@@

\paragraph{Step6}\hfill \\ 
Step 6 requires a \gls{json} parsing of the decoded `cData'. Step 5 and 6 are to close together that the reference implementation combined them into one step which deals with both at the same time. The final clientDataJSON can is displayed in Listing \ref{lst:attestationClientData}.

@@

\paragraph{Step9}\hfill \\ 
In step 9 the origin has to be verified. Again as in the registration process, this is done by checking if the origin contains the domain. In a productive environment, this should be done by comparing the expected URI.

\paragraph{Step10}\hfill \\ 
Step 10 deals with the token binding over \gls{tls} which is optional and currently not supported in the reference implementation.

\paragraph{Step11}\hfill \\ 
In step 11 the rpIdHash from the `authData' must equal to the \gls{sha-256} hash of the RP ID. For this first, the `authData' needs to be parsed.

\paragraph{Step12}\hfill \\ 
In step 12 the user present flag is checked. This check is the same process as in the registration and must be verified to protect against remote attacks.

@@

\paragraph{Step15}\hfill \\ 
In step 15 the client data is hashed using \gls{sha-256}. The data provided needs to be the exact string that was received by accessing the clientDataJSON field.

@@

\paragraph{Step17}\hfill \\ 
Step 17 is needed to check for cloned authenticators. If the sign count is present, it should always increase. It is important to verify this behavior to avoid cloned authenticators.

\paragraph{Step18}\hfill \\ 
Step 18 states that should any step above have been invalid; the authentication ceremony must be failed.

\subsubsection{Summary}
After already knowing what happens in the registration the authentication is very similar as many steps exist in both ceremonies. The steps of the assertion sometimes can't be represented correctly in code because with java it is sometimes easier to combine two steps into one. However, it is important that all of the steps are followed as required. Missing some steps could lead to broken authentication.

\subsection{Auth Data}
\label{sec:authData}
In both the registration and authentication ceremony, there is one step where the authenticator data must be parsed. Thus this was done at a central place called the `AuthenticatorDataParser'. This data comes in binary and needs to be parsed according to the specification. The first 32 bytes generate the rpIdHash, the next byte the flags sector. Then there are 4 bytes which represent the signature Counter. After that, there might exist some attestedCredentialData. A table with all the information can be seen in Figure \ref{fig:authenticatorData}.

The attestedCredentialData if present contains 16 bytes of the `aaguid', with which the authenticator model can be determined, 2 bytes which set the length of the following field which is the credential id and the rest which makes the credential Public Key. This attestedCredentialData is present for registration to verify the assertion. A table with all the information can be seen in Figure \ref{fig:credentialData}.

@@

\section{Conclusion}
The \gls{fido2} standard increases the security of the end user by eliminating passwords with a \gls{mitm} resistant factor. The authenticator can be supplied by many parties which makes wide adoption easier. However, currently, the implementation of the standard for a webpage owner is probably to complex. There are not yet convenient libraries for all languages available. This needs to change if wide adoption is to be achieved. Additionally, there are not yet many \gls{fido2} authenticators available on the market. Which results in an even poorer adoption rate.

However, the standard is considerably new and could be adopted widely. It has strong support from Microsoft, where it is implemented in Windows Hello, and other big technology companies. The implementation in Windows Hello could result in many companies supplying authenticators to their employees which would lead to a higher incentive to provide web applications with \gls{fido2} support. It is crucial that other big players indeed implement the standard for core products so that it truly can replace passwords.

An exciting application of this would be within companies to replace the need for passwords. Also, it could be used by governments in combination with an \gls{eid} to give access to specific resources by using a physical card as an authenticator. This way every citizen of a country already has access to a \gls{fido} certified authenticator which would lead to country-specific sites implementing the standard as an authentication method.

@@

This implementation only covered the Yubikey 5 authenticator. It would be interesting to see how other authenticators react. Also, it would be interesting to see how the Yubikey 5 interacts via the \gls{nfc} interface as this implementation was only tested using the \gls{usb} interface of the authenticator.

Additionally, it only dealt with authentication and registration. It is essential to give the user the option to add multiple authenticators to his account such that he can recover his account should he lose his key. Additionally, it is vital that the user can revoke their authenticators. This possibility should be done by giving the user the option to set a name that is relevant to him for each authenticator. He should then be able to manage those authenticators. Either adding new ones or removing existing ones.
